---
title: "Cyclistic_Capstone1"
author: "Mostafa"
date: "2023-07-29"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
options(repos = "https://cran.rstudio.com/")  # Set the CRAN mirror URL

```



## Preparing Phase | Setting up my enviroment 
### Installing the packages
```{r Installing the packages, message=FALSE, warning=FALSE}
install.packages("tidyverse")
install.packages("readr")
install.packages("janitor")
install.packages("tidyr")
install.packages("dplyr")
install.packages("ggplot2")

```



### Loading the packages
```{r}
library(tidyverse)
library(readr)
library(janitor)
library(tidyr)
library(dplyr)
library(ggplot2)

```



### Importing & naming the datasets
```{r}
Jan2021= read_csv("CyclisticTripdata202101.csv")
Feb2021= read_csv("CyclisticTripdata202102.csv")
Mar2021= read_csv("CyclisticTripdata202103.csv")
Apr2021= read_csv("CyclisticTripdata202104.csv")
May2021= read_csv("CyclisticTripdata202105.csv")
June2021= read_csv("CyclisticTripdata202106.csv")
July2021= read_csv("CyclisticTripdata202107.csv")
Aug2021= read_csv("CyclisticTripdata202108.csv")
Sep2021= read_csv("CyclisticTripdata202109.csv")
Oct2021= read_csv("CyclisticTripdata202110.csv")
Nov2021= read_csv("CyclisticTripdata202111.csv")
Dec2021= read_csv("CyclisticTripdata202112.csv")
```



## Process Phase | Cleaning the datasets

### Check the data type
```{r}
str(Jan2021)
str(Feb2021)
str(Mar2021)
str(Apr2021)
str(May2021)
str(June2021)
str(July2021)
str(Aug2021)
str(Sep2021)
str(Oct2021)
str(Nov2021)
str(Dec2021)
```



### Creating the dataframe
```{r}
MonthlyTrip_df = bind_rows(Jan2021,Feb2021, Mar2021, Apr2021,May2021,June2021,July2021,Aug2021,Sep2021,Oct2021,Nov2021,Dec2021)
```



### Cleaning columns names
```{r}
MonthlyTrip_df = clean_names(MonthlyTrip_df)
```



### Removing all the full empty columns and rows
```{r}
MonthlyTrip_df = remove_empty(MonthlyTrip_df, which = c())
```



### Removing duplicates
```{r}
MonthlyTrip_df = unique(MonthlyTrip_df)
```



### Check on the set of values in member_casual field
```{r}
all(MonthlyTrip_df$member_casual %in% c("member","casual"))

```


### Removing leading & trailing spaces
```{r}
MonthlyTrip_df %>% 
  mutate_all(trimws)
```


### Removing in between excessive spaces
```{r}
MonthlyTrip_df %>% 
  mutate_all(~gsub("\\s+"," ", .))
```



To answer some of the business tasks I need to extract the following:

* Weekdays
* Hours
* Months

And  in the same context, calculate the tripduration

### Extracting weekday with wday() and creating day_of_week new column
```{r}
MonthlyTrip_df$day_of_week= wday(MonthlyTrip_df$started_at, label = T, abbr = F)
```




### Extracting Time-Hour
```{r}
MonthlyTrip_df$starting_hour = format(as.POSIXct(MonthlyTrip_df$started_at),"%H")

```



### Extracting Month
```{r}
MonthlyTrip_df$starting_month = format(as.Date(MonthlyTrip_df$started_at), "%m")

```




### Calculate tripduration
```{r}
MonthlyTrip_df$tripduration = difftime(MonthlyTrip_df$ended_at,MonthlyTrip_df$started_at,units= "min")
```


Converting "tripduration" from difftime to numeric
```{r}
MonthlyTrip_df$tripduration = as.numeric(MonthlyTrip_df$tripduration)

```



Removing data with tripduration =< 0
```{r}
MonthlyTrip_df = MonthlyTrip_df[!(MonthlyTrip_df$tripduration<=0),]

```


### tripduration outliers
```{r}
boxplot(MonthlyTrip_df$tripduration)
q1 = quantile(MonthlyTrip_df$tripduration,0.25)
q3= quantile(MonthlyTrip_df$tripduration,0.75)
iqr = IQR(MonthlyTrip_df$tripduration)
Cleaned_MonthlyTrip_df = MonthlyTrip_df %>% 
  filter(!(MonthlyTrip_df$tripduration> q3 +1.5*iqr | MonthlyTrip_df$tripduration< q1 -1.5*iqr))
```


  
Tripduration statistics summary before filtration
```{r}
summary(MonthlyTrip_df$tripduration)
```
Tripduration statistics summary before filtration
```{r}
summary(Cleaned_MonthlyTrip_df$tripduration)
```






*Creating cleaned csv data file to use it later in Tableau*
```{r}
write.csv(Cleaned_MonthlyTrip_df, file= "Cyclistic_trips.csv")
```






## Analysis and Visulizations

**For the business tasks I need analyze & visualize the following to distinguish differences between the users**

###  Number of Rides for User Type in Weekdays
```{r}
ggplot(data = Cleaned_MonthlyTrip_df) + 
  geom_bar(mapping = aes(x= day_of_week, fill = member_casual), position = "dodge") +
  labs(x = "Weekday", y = "Number of Rides", fill = "Member Type", title = "Number of Rides for Member Type in Weekdays")+
  theme(axis.text = element_text(size = 6))
ggsave("number_of_rides_for_member_type_weekdays.png")
```





### Number of Rides for Member Type in Months
```{r}
options(scipen = 1000)
ggplot(data = Cleaned_MonthlyTrip_df) + 
  geom_bar(mapping = aes(x= starting_month, fill = member_casual), position = "dodge") +
  labs(x = "Month", y = "Number of Rides", fill = "Member Type", title = "Number of Rides for Member Type in Months") +
  theme(axis.text = element_text(size = 7))
ggsave("number_of_rides_for_member_type_months.png")

```




### Number of Rides for Member Type per daily hrs
```{r}
options(scipen = 1000)
ggplot(data = Cleaned_MonthlyTrip_df) + 
  geom_bar(mapping = aes(x= starting_hour, fill = member_casual), position = "dodge") +
  labs(x = "Hour", y = "Number of Rides", fill = "Member Type", title = "Number of Rides for Member Type in Daily Hours") +
  theme(axis.text = element_text(size = 7)) + 
  facet_wrap(~day_of_week)
ggsave("number_of_rides_for_member_type_hours.png")
```





### The average trip duration within weekdays viz between casuals & members
```{r}
average_tripduration = Cleaned_MonthlyTrip_df %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise(avg_tripduration= mean(tripduration))

options(scipen = 1000)
ggplot(data = average_tripduration) + 
  geom_col(mapping = aes(x=day_of_week, y= avg_tripduration, fill = member_casual), position = "dodge") +
  labs(x = "Weekday", y = "AVG Trip Duration in sec.", fill = "Member Type", title = "Average Trip Duration in Seconds") +
  theme(axis.text = element_text(size = 7))
ggsave("avg_tripduration_seconds.png")
```




### The top 10 stations for the number of trips for members
```{r message=FALSE, warning=FALSE}
Cleaned_MonthlyTrip_df %>%
  filter(member_casual=="member" & !is.na(start_station_name)) %>% 
  count(start_station_name) %>%
  top_n(10) %>%
  arrange(desc(n)) %>%
  ggplot() +
  geom_col(mapping = aes(y = reorder(start_station_name, n), x = n)) +
  geom_text(mapping = aes(y = reorder(start_station_name, n), x = n, label = n), vjust = -0.5) +
  labs(x = "Count") +
  theme(axis.text = element_text(size = 7))
```


  
### The top 10 stations for the number of trips for casuals
```{r}
Cleaned_MonthlyTrip_df %>%
  filter(member_casual=="casual" & !is.na(start_station_name)) %>% 
  count(start_station_name) %>%
  top_n(10) %>%
  arrange(desc(n)) %>%
  ggplot() +
  geom_col(mapping = aes(y = reorder(start_station_name, n), x = n)) +
  geom_text(mapping = aes(y = reorder(start_station_name, n), x = n, label = n), vjust = -0.5) +
  labs(x = "Count") +
  theme(axis.text = element_text(size = 7))
ggsave("top_10_stations_casual.png")

```


